---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![Build Status](https://travis-ci.org/RobinHankin/permutations.svg?branch=master)](https://travis-ci.org/RobinHankin/permutations)
<!-- badges: end -->


# Overview

The permutations package provides R-centric functionality for working
with .  It includes cross products and wedge products and a variety of
use-cases.  An extensive vignette is provided in the package; this
document is here for discoverability.


# Installation

You can install the released version of permutations from [CRAN](https://CRAN.R-project.org) with:

```{r}
# install.packages("permutations")  # uncomment this to use the package
library("permutations")
```

The package is maintained on
[github](https://github.com/RobinHankin/permutations).	

# The ```permutations``` package in use

Random permutations on a finite set are given by the ```rperm()``` command:

```{r rperm}
rperm(10,9)
```

The result is printed in cycle form but we can print in word form if we wish:

```{r wordformprint}
options(print_word_as_cycle=FALSE)  # override default
as.word(rperm(10,9,3))
options(print_word_as_cycle=TRUE)  # usually want to print a cycle irregardless
```

(A dot indicates a fixed point).  The package uses arithmetic
operations ```*``` to combine permutations and ```^``` for
conjugation:

```{r arith}
(a <- as.word(c(4,2,3,1,5,7,6)))
(b <- as.cycle(1:4))
a*b
b*a
```

# The megaminx

The *megaminx* is a dodecahedral puzzle with similar construction to
the Rubik cube puzzle that has 50 movable pieces and 132 coloured
stickers (``facets'').  The permutations package includes
functionality to simulate the megaminx and exhibits an 82-turn
superflip.  The vignette gives an extended discussion.
