---
title: "Enumerating all partitions of a given shape"
output: html_document
date: "2023-06-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TLDR summary

$\ldots$ I couldn't do it. But I could count them!

## Combinatorics

How many permutations are there of a given shape? If the shape is

$$
\underbrace{r_1,\ldots,r_1}_{m_1},
\underbrace{r_2,\ldots,r_2}_{m_2},\ldots,
\underbrace{r_d,\ldots,r_d}_{m_d}
$$

(that is, there are $m_1$ cycles of size $r_1$, $m_2$ cycles of size
$r_2$, etc; we understand that $r_1>r_2>\cdots>r_d$). Writing
$\sum_{i=1}^d m_ir_i=n$ for the size of the permuted set [NB $r_d=1$
is permissible] then I get

$${n\choose 
\underbrace{r_1,\ldots,r_1}_{m_1},
\underbrace{r_2,\ldots,r_2}_{m_2},\ldots,
\underbrace{r_d,\ldots,r_d}_{m_d}
}\prod_{i=1}^d\frac{\Gamma(r_i)^{m_i}}{m_i!}
$$

[stackexchange](https://math.stackexchange.com/questions/1323130/how-many-permutations-of-cycle-shape-3-22-1-are-there-in-s-8)
has a good question here, the answer agrees with mine. He asks how
many permutations of shape $3,2,2,1$, that is,
$(\cdot\cdot\cdot)(\cdot\,\cdot)(\cdot\,\cdot)(\cdot)$ and gets 1680.
I get:

$$
{8\choose 3\,2\,2\,1}\frac{2}{2!}=5678=30\cdot 56=1680
$$

We might be able to make a start on enumerating this by using the
`partitions` package:

```{r}
library("partitions")
M <- multinomial(c(a=3,b=2,c=2,d=1))
summary(M)
```

Long experience has shown me that taking the first column can be
non-intuitive and confusing. So I will take the last column as an
example. The last column, viz `67845231` corresponds to permutation
$(678)(45)(23)(1)$. Now on the one hand we somehow need to flag that
$$(678)(45)(23)(1)=(678)(23)(45)(1)$$ [i.e. presenting the 2-cycles in
a different order does not change the permutation: boxes `b` and `c`
are indistinguishable]; and
$$(678)(45)(23)(1)=(786)(45)(23)(1)=(678)(54)(23)(1),$$ but also to
note that $$(678)(45)(23)(1)\neq(687)(45)(23)(1).$$ To be explicit,
`multinomial()` will include $(678)(45)(23)(1)$ and $(678)(23)(45)(1)$
[thus over-counting by a factor of two] but omit $(687)(23)(45)(1)$
[thus under-counting by a factor of two].

Not sure of the best way to proceed: perhaps
`partitions::multinomial()` could have a flag that specifies whether
boxes with the same capacities are indistinguishble (currently they
are not):

```{r}
multinomial(c(a=2,b=2))
```

Above we see $(12)(34)$ and $(34)(12)$, which are the same partition
if boxes `a` and `b` are indistinguishable.